Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2022-02-26 unimodular-covers.tex
edited ref to Gubeladze and updated bibliography

\usepackage{amssymb}
\usepackage{graphicx,color}
\usepackage{amsmath}
\usepackage{url}
\usepackage[colorlinks=true]{hyperref}
%\usepackage{pictex}

@@

The two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP} is known to hold, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}. This last one actually shows that smoothness of $Q$ is not needed. In dimension three, however, the conjecture fails without the smoothness assumption. Indeed, if we let $P=Q$ be any non-unimodular \emph{empty tetrahedron}, then $P$ is obviously a weak Minkowski summand of $Q$ but the pair $(P.Q)$ is not IDP. By an empty tetrahedron we mean a lattice tetrahedron containing no lattice points other than its vertices (see the proof of \Cref{lemma:corner} for a classification of them).

@@

As we note in \Cref{prop:mixedIDP}, if the Cayley sum of $P$ and $Q$ is IDP then the pair $(P,Q)$ is IDP.
In particular, the following statement from \Cref{sec:cayley} is stronger than the afore-mentioned result of \cite{Fakhruddin,HNPS2008,Ogata}:

@@

This has the following consequence, also proved in \Cref{sec:cayley}.
A  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being ($SL(\Z,d)$-equivalent to) a Cayley sum.

@@

\paco{deleted former corollary}
%\Paco{
%I am a bit puzzled. After the next corollary we say that \cite[Cor.~4.2]{SantosZiegler} proves a case of this, but in fact it proves the whole of it, in a stronger form: it does not need width one, and it gives a "dissection" and not a mere cover (dissection= cover by simplices with disjoint interiors). 
%I propose to delete it:
%}
%{\color{red}
%\begin{corollary}
%%\label{coro:width1}
%Every integer dilation $kP$, $k\ge 2$, of a lattice $3$-polytope $P$ of width $1$ has a unimodular cover.
%\end{corollary}
%
%A special case of the latter are integer dilations of empty tetrahedra. That their dilations have
%unimodular covers is \cite[Cor.~4.2]{SantosZiegler} (and is also implicit in \cite{KantorSarkaria}).
%}

\paco{added ref to Gubeladze}
\giulia{changed it}

To date there are no known example of IDP $3$-polytopes without a unimodular cover. Recent work of Gubeladze~\cite{Gubeladze} shows that another class of IDP 3-polytopes has unimodular covers, lattice polytopes which are convex hulls of all lattice points inside an ellipsoid. It remains open in general whether the classes are distinct in dimension $3$. 
%Let us also mention the following recent results of Gubeladze~\cite{Gubeladze}.
%In part (2), a polytope $P$ is called \emph{very ample} if the vertex cone $\cone(P-v)$ is contained in the translation $P-v$, for every vertex $v$ of $P$.

@@

\paco{removed closing sentence}
%\medskip
%We believe that the $3$-polytopes in all these statements have unimodular triangulations, but this remains an open question.

@@

Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. Then:
\[
w=\width_{f_q}(p + \vecline q )=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\vecline p ). 
\]

@@

\bibitem{BHHHJKM2019}
Matthias Beck, Christian Haase, Akihiro Higashitani, Johannes Hofscheier, Katharina Jochemko, Lukas Katth\"an, Mateusz Micha{\l}ek,
Smooth Centrally Symmetric Polytopes in Dimension 3 are IDP,
\emph{Ann.~Combin.}, in press, 2019.
\url{https://doi.org/10.1007/s00026-019-00418-x}

@@

@@

\bibitem{Gubeladze}
Joseph Gubeladze,
Normal polytopes and ellipsoids,
preprint, \href{https://arxiv.org/abs/2012.11846}{arXiv:2012.11846}, December 2012.

\bibitem{HaaseHof}
Christian Haase and Jan Hofmann, 
Convex-normal (Pairs of) Polytopes, 
\emph{Canad. Math. Bull.} 60 (2017), 510--521.

  \bibitem{HPPS-survey}
Christian Haase, Andreas Paffenholz, Lindsay C. Piechnik, Francisco Santos. Existence of unimodular triangulations - positive results. 
%Preprint May 2014, updated December 2017, 89 pages. 
 \emph{Mem. Amer. Math. Soc.}, to appear.
 Available as arXiv preprint \href{https://arxiv.org/abs/1405.1687}{arXiv:1405.16878}

\bibitem{HNPS2008}
Christian Haase, Benjamin Nill, Andreas Paffenholz, and Francisco Santos, Lattice points in Minkowski sums, 
\emph{Electron.~J.~Combin.}, 15 (2008), no. 1, Note 11, 5 pp.

\bibitem{KantorSarkaria}
Jean-Michel Kantor and Karanbir~S.\ Sarkaria.
On primitive subdivisions of an elementary tetrahedron, 
\emph{Pacific J.~Math.} \textbf{211} (2003), no.~1,  123--155.

\bibitem{mfo2004}
Mini-workshop: Ehrhart Quasipolynomials: Algebra, Combinatorics, and Geometry, Oberwolfach Rep. 1 (2004), no. 3, 2071-2101, Abstracts from the mini-workshop held August 15-21, 2004, Organized by Jes\'us A.~De Loera and Christian Haase, Oberwolfach Reports. Vol. 1, no. 3. MR MR2144157

\bibitem{mfo2007}
Mini-workshop: Projective normality of smooth toric varieties, Oberwolfach Rep. 4 (2007), no. 39/2007, Abstracts from the mini-workshop held August 12-18, 2007. Organized by Christian Haase, Takayuki Hibi and Diane MacLagan.

@@

\bibitem{Ogata}
Shoetsu Ogata.
\newblock {Multiplication maps of complete linear systems on projective toric surfaces.}
\newblock {\em Interdiscip. Inf. Sci.}, 12(2):93--107, 2006.

\bibitem{Santos-hirsch}
Francisco Santos.
A counter-example to the Hirsch Conjecture.
\emph{Ann.~Math.~(2)}, 176 (July 2012), 383--412. 
DOI: 10.4007/annals.2012.176.1.7

\bibitem{SantosZiegler}
Francisco Santos and G{\"u}nter~M. Ziegler.
\newblock {Unimodular triangulations of dilated 3-polytopes}, 
\newblock {\em Trans. Moscow Math. Soc.} (2013), 293--311. % \MR{3235802}

@@

\bibitem{White1964}
G.~K.~White.
Lattice tetrahedra.
\emph{Canadian J.~Math.} 16 (1964), 389--396.

>>>>

\usepackage{amssymb}
\usepackage{graphicx,color}
\usepackage{amsmath}
\usepackage{url}
\usepackage[colorlinks=true]{hyperref}
\usepackage{doi}
%\usepackage{pictex}

@@

The two-dimensional case of \Cref{conj:Oda}\eqref{itm:mixedIDP} is known to hold, with three different proofs by Fakhruddin~\cite{Fakhruddin}, Ogata~\cite{Ogata} and Haase et al.~\cite{HNPS2008}. This last one actually shows that smoothness of $Q$ is not needed. In dimension three, however, the conjecture fails without the smoothness assumption. Indeed, if we let $P=Q$ be any non-unimodular \emph{empty tetrahedron}, then $P$ is obviously a weak Minkowski summand of $Q$ but the pair $(P, Q)$ is not IDP. By an empty tetrahedron we mean a lattice tetrahedron containing no lattice points other than its vertices (see the proof of \Cref{lemma:corner} for a classification of them).

@@

As we note in \Cref{prop:mixedIDP}, if the Cayley sum of $P$ and $Q$ is IDP then the pair $(P,Q)$ is IDP.
Hence, the following statement proved in \Cref{sec:cayley} is stronger than the afore-mentioned result of \cite{Fakhruddin,HNPS2008,Ogata}:

@@

This has the following consequence, also proved in \Cref{sec:cayley}.
Here a  \emph{prismatoid} is a polytope whose vertices all lie in two parallel facets. 
%A polytope has width $1$ if its vertices lie in two \emph{consecutive} parallel lattice hyperplanes. Observe that this is the same as being ($SL(\Z,d)$-equivalent to) a Cayley sum.

@@

%\paco{deleted former corollary}
%\Paco{
%I am a bit puzzled. After the next corollary we say that \cite[Cor.~4.2]{SantosZiegler} proves a case of this, but in fact it proves the whole of it, in a stronger form: it does not need width one, and it gives a "dissection" and not a mere cover (dissection= cover by simplices with disjoint interiors). 
%I propose to delete it:
%}
%{\color{red}
%\begin{corollary}
%%\label{coro:width1}
%Every integer dilation $kP$, $k\ge 2$, of a lattice $3$-polytope $P$ of width $1$ has a unimodular cover.
%\end{corollary}
%
%A special case of the latter are integer dilations of empty tetrahedra. That their dilations have
%unimodular covers is \cite[Cor.~4.2]{SantosZiegler} (and is also implicit in \cite{KantorSarkaria}).
%}

\paco{added ref to Gubeladze}
\giulia{changed it}
\paco{changed again, adding more references}

Let us mention that recent work of Gubeladze~\cite{Gubeladze} shows another class of 3-polytopes admitting unimodular covers: the convex hulls of all lattice points inside an ellipsoid; these had previously been shown to be IDP by Bruns, Gubeladze and Michalek~\cite{BGM}.
To date there are no known examples of IDP $3$-polytopes without a unimodular cover, although such polytopes exist in higher dimension~\cite{BG}.
%Let us also mention the following recent results of Gubeladze~\cite{Gubeladze}.
%In part (2), a polytope $P$ is called \emph{very ample} if the vertex cone $\cone(P-v)$ is contained in the translation $P-v$, for every vertex $v$ of $P$.

@@

%\paco{removed closing sentence}
%\medskip
%We believe that the $3$-polytopes in all these statements have unimodular triangulations, but this remains an open question.

@@

Let $w=\area(p+q) \geq 2$, where $\area$ denotes the area normalized to a fundamental domain. In what follows, the width of a functional $f$ on a set $S$, denoted $\width_f(S)$ is defined as the difference $\sup_{x\in S} f(x)-\inf_{x\in S} f(x)$
Then:
\[
w=\width_{f_q}(p + \vecline q )=\width_{f_q}(p)=\width_{f_p}(q)=\width_{f_p}(q +\vecline p ). 
\]

@@

\bibitem{BHHHJKM2019}
Matthias Beck, Christian Haase, Akihiro Higashitani, Johannes Hofscheier, Katharina Jochemko, Lukas Katth\"an, Mateusz Micha{\l}ek,
Smooth Centrally Symmetric Polytopes in Dimension 3 are IDP,
\emph{Ann.~Combin.}, 23 (2019), 255--262.
\doi{10.1007/s00026-019-00418-x}

\bibitem{BG}
Winfried Bruns and Joseph Gubeladze, Normality and covering properties of affine semigroups, \emph{J. Reine Angew. Math.} 510 (1999), 151--178. \doi{10.1515/crll.1999.044}

@@

\bibitem{BGM}
  Winfried Bruns, Joseph Gubeladze, and Mateusz Michalek. Quantum Jumps of Normal Polytopes. \emph{Discrete Comput. Geom.}, 56(1) (2016), 181--215. \doi{10.1007/s00454-016-9773-7}

@@

\bibitem{Gubeladze}
Joseph Gubeladze,
Normal polytopes and ellipsoids,
\emph{Elect. J. Combin.}, 28:4 (2021), P4.7.
\doi{10.37236/10338}

\bibitem{HaaseHof}
Christian Haase and Jan Hofmann, 
Convex-normal (Pairs of) Polytopes, 
\emph{Canad. Math. Bull.} 60:3 (2017), 510--521. \doi{10.4153/CMB-2016-057-0}

  \bibitem{HPPS-survey}
Christian Haase, Andreas Paffenholz, Lindsay C. Piechnik, Francisco Santos. Existence of unimodular triangulations - positive results. 
 \emph{Mem. Amer. Math. Soc.},  Vol.~270 1N.~321, American Math. Society, 2021. DOI: \doi{10.1090/memo/1321}
%ISBNs: 978-1-4704-4716-8 (print); 978-1-4704-6530-8 (online)

\bibitem{HNPS2008}
Christian Haase, Benjamin Nill, Andreas Paffenholz, and Francisco Santos, Lattice points in Minkowski sums, 
\emph{Electron.~J.~Combin.}, 15:1 (2008), Note 11, 5 pp. \doi{10.37236/886}

\bibitem{KantorSarkaria}
Jean-Michel Kantor and Karanbir~S.\ Sarkaria.
On primitive subdivisions of an elementary tetrahedron, 
\emph{Pacific J.~Math.} \textbf{211} (2003), no.~1,  123--155. \doi{10.2140/pjm.2003.211.123}

\bibitem{mfo2004}
Mini-workshop: Ehrhart Quasipolynomials: Algebra, Combinatorics, and Geometry, Oberwolfach Rep. 1 (2004), no. 3, 2071-2101, Abstracts from the mini-workshop held August 15-21, 2004, Organized by Jes\'us A.~De Loera and Christian Haase, Oberwolfach Reports. Vol. 1, no. 3. \doi{10.14760/OWR-2004-39}

\bibitem{mfo2007}
Mini-workshop: Projective normality of smooth toric varieties, Oberwolfach Rep. 4 (2007), no. 39/2007, Abstracts from the mini-workshop held August 12-18, 2007. Organized by Christian Haase, Takayuki Hibi and Diane MacLagan. \doi{10.14760/OWR-2007-39}

@@

\bibitem{Ogata}
Shoetsu Ogata.
\newblock {Multiplication maps of complete linear systems on projective toric surfaces.}
\newblock {\em Interdiscip. Inf. Sci.}, 12(2):93--107, 2006. \doi{10.4036/iis.2006.93}

\bibitem{Santos-hirsch}
Francisco Santos.
A counter-example to the Hirsch Conjecture.
\emph{Ann.~Math.~(2)}, 176 (July 2012), 383--412. 
\doi{10.4007/annals.2012.176.1.7}

\bibitem{SantosZiegler}
Francisco Santos and G{\"u}nter~M. Ziegler.
\newblock {Unimodular triangulations of dilated 3-polytopes}, 
\newblock {\em Trans. Moscow Math. Soc.} (2013), 293--311. \doi{10.1007/s00493-014-2959-9}

@@

\bibitem{White1964}
G.~K.~White.
Lattice tetrahedra.
\emph{Canadian J.~Math.} 16 (1964), 389--396. \doi{10.4153/CJM-1964-040-2}
